---
title: "Mapping crash data"
output:
  html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using the processed shapefile of towns in the six-county area around Chicago, we can explore the pedestrian/cycling crash data using choropleth maps. 

A choropleth map uses color to identify different categories of a set of information. The most recognizable choropleth are ones used for elections, showing states that voted GOP (red) or Democratic (blue).

## Loading data, map styles

```{r, echo=FALSE}

# check libraries
library(readr)
library(rgdal)
library(dplyr)
library(RColorBrewer) # for my color ramp
library(viridis)
library(scales)
library(ggplot2)
library(ggmap)
library(ggthemes)

#---------------------
# This function set styles for the chart
# Be sure to run it before you plot

theme_map <- function(...) {
  theme_set(theme_get() + theme(text = element_text(family = 'Verdana', size= 12, lineheight=0.9))) + 
    theme(
      # edit background colors and elements
      plot.background = element_blank(),
      legend.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      axis.ticks = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank(),
      # edit font sizes
      plot.title = element_text(size = 27,face="bold"),
      plot.subtitle = element_text(size = 18),
      legend.title=element_text(size = 13,face="bold"),
      legend.text=element_text(size=14.7),
      legend.text.align = 0, # 0 = left, 1 = right
      plot.caption=element_text(size=13.5, hjust=0),
      strip.text = element_text(face="bold", size=13.5, hjust=0),
       ...
    )
}

#-----------------
# read in the data

ill_f <- read.csv("data/ill_f.csv", stringsAsFactors = FALSE)

co6 <- read.csv("data/counties6.csv", stringsAsFactors = FALSE)


```

##How we looked at the data geographically

Using the R code included, we created several maps to help us examine if there were geographic patterns hidden in the numbers. 

We mapped total crashes by year and crashes with injuries by year. We also  calculated the rate of crashes and the rate of injuries for each municipality by 1,000 population, for each year. 

```{r  eval = FALSE}

ill_f$total_injrate_2014 <- (ill_f$total_inj_2014/ill_f$pop_2014)*1000
ill_f$total_injrate_2015 <- (ill_f$total_inj_2015/ill_f$pop_2015)*1000
ill_f$totalinjchg_1415 <- ill_f$total_injrate_2015 - ill_f$total_injrate_2014

```

Then we started generating maps. Below are some examples.

##2015 crashes

Here's the rate per 1000 population for all crashes in 2015, by town. The data is mapped by quintiles, or in five chunks with about the same number of towns in each chunk.

Here's how that would divide up.

```{r}
quantile(ill_f$total_rate_2015, probs = seq(0, 1, .2), na.rm=TRUE)
```

This shows the break points in our data for quintiles. Each chunk is about 20% of the data (100/5 = 20). The lowest 20%, from 0 to 20%, is 0.04 to 0.18. The highest 20% is from 0.49 to 4.81. 

Here's what that looks like on the map.

```{r, fig.width=10, fig.height=10}


#----------------------------------
#----------------------------------

# define quantiles for data
no_classes <- 5
labels <- c()

quantiles <- quantile(ill_f$total_rate_2015, # data you wish to work with
                      probs = seq(0, 1, length.out = no_classes + 1),
                      na.rm=TRUE)

# custom labels
labels <- c()
for(idx in 1:length(quantiles)){
  labels <- c(labels, paste0(round(quantiles[idx], 2)))
}
labels <- labels[1:length(labels)-1]

# add quantile info to dataset
ill_f$total_rate_2015_quantiles <- cut(ill_f$total_rate_2015, 
                                   breaks = quantiles, 
                                   labels = labels, 
                                   include.lowest = T)

# now map

map <- ggplot(ill_f, 
              aes(long, lat, group = group, fill = total_rate_2015_quantiles)
) +
  geom_polygon( color = "white", size = 0.15) + coord_map() + theme_map()

map <- map + labs(
  fill = "2015 crash rate"
  )

map <- map + 
  theme(legend.position = "top", legend.direction = "horizontal") +
  scale_fill_manual(
  values = c("#a6611a", "#dfc27d", "#f6e8c3", "#80cdc1", "#018571"),
  guide = guide_legend(
    direction = "horizontal",
    keyheight = unit(2, units = "mm"),
    keywidth = unit(75 / length(labels), units = "mm"),
    title.position = 'top',
    title.hjust = 0.5,
    nrow = 1,
    byrow = T,
    # Can reverse the legend if that's clearer
    reverse = F,
    label.position = "bottom")
)

map <- map + geom_polygon(data = co6, aes(x=long, y = lat, group = group),
                          fill = NA, color = "blue", size = 0.25)

map

```

##Change over time

We looked how the data changed over time. For instance, here's the change in the total crash rate per 1000 from 2014 to 2015.


```{r, fig.width=10, fig.height=10}

ill_f$chg1415 <- ill_f$total_rate_2015 - ill_f$total_rate_2014
quantile(ill_f$chg1415, probs = seq(0, 1, .2), na.rm=TRUE)

# define quantiles for data
no_classes <- 5
labels <- c()

quantiles <- quantile(ill_f$chg1415, # data you wish to work with
                      probs = seq(0, 1, length.out = no_classes + 1),
                      na.rm=TRUE)

# custom labels
labels <- c()
for(idx in 1:length(quantiles)){
  labels <- c(labels, paste0(round(quantiles[idx], 2)))
}
labels <- labels[1:length(labels)-1]

# add quantile info to dataset
ill_f$chg1415_quantiles <- cut(ill_f$chg1415, 
                                       breaks = quantiles, 
                                       labels = labels, 
                                       include.lowest = T)

# now map

map <- ggplot(ill_f, 
              aes(long, lat, group = group, fill = chg1415_quantiles)
) +
  geom_polygon( color = "white", size = 0.15) + coord_map() + theme_map()

map <- map + labs(
  fill = "Change in total rate, 2014-2015"
)

map <- map + 
  theme(legend.position = "top", legend.direction = "horizontal") +
  scale_fill_manual(
    values = c("#a6611a", "#dfc27d", "#f6e8c3", "#80cdc1", "#018571"),
    guide = guide_legend(
      direction = "horizontal",
      keyheight = unit(2, units = "mm"),
      keywidth = unit(75 / length(labels), units = "mm"),
      title.position = 'top',
      title.hjust = 0.5,
      #label.hjust = 1,
      nrow = 1,
      byrow = T,
      # Can reverse the legend if that's clearer
      reverse = F,
      label.position = "bottom")
  )

map <- map + geom_polygon(data = co6, aes(x=long, y = lat, group = group),
                          fill = NA, color = "blue", size = 0.25)

map
```

The biggest problem with these maps is that it's capturing places with relatively low population but with high rates. Prime example being Barringtion, located at the point were Lake, Cook, Kane and McHenry counties touch.

##Mapping totals

In the end, we found the most compelling and useful map was one showing how pervasive the problem is.

We mapped towns with five or more total crashes with fatalities over the four years (dark green), towns with at least one and fewer than five crashes with fatalities (light green) and then towns with only crashes with injuries (brown).

```{r, fig.width=10, fig.height=10}

towns_f <- read.csv("data/towns_f.csv")

fatals <- towns_f[towns_f$total_fatal > 0,]
fatals5 <- towns_f[towns_f$total_fatal >= 5,]
injured <- towns_f[towns_f$total_inj > 0,]

map <- ggplot() + 
  geom_polygon(data = towns_f, aes(long, lat, group = group), 
               fill = "#f6e8c3", color = "white", size = 0.15) + coord_map() + theme_map()

map <- map + geom_polygon(data = injured, aes(long, lat, group = group), 
               fill = "#dfc27d", color = "white", size = 0.15) + coord_map()

map <- map + geom_polygon(data = fatals, aes(long, lat, group = group), 
               fill = "#80cdc1", color = "white", size = 0.15) + coord_map()

map <- map + geom_polygon(data = fatals5, aes(long, lat, group = group), 
               fill = "#018571", color = "white", size = 0.15) + coord_map()

map <- map + theme(legend.position="None")

map <- map + geom_polygon(data = co6, aes(x=long, y = lat, group = group),
                          fill = NA, color = "blue", size = 0.25)

map

```

Most of the areas in dark green are places with high populations, and so we expect to see them.
But about five places - including Des Plaines and Skokie - had more fatalities in the time examined but were not among places with the highest population.
Finally, we exported out the data to use as the basis of a web app, and to import into QGis to create the geojson that would power an interactive map. 